The Ranked Nearest Neighbour Interchange space

### 2019-05-01 paper.tex
update proof convexity of caterpillars with shared cherry taxon

@@

\begin{proof}
	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
	Let $p$ be a shortest path from $T_1$ to $T_2$.
	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.

@@

	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
	Let $A$ and $B$ be the subtrees that are children of $int_x$ and $C$ the subtree that is child of $int_y$ and does not contain $A$ and $B$, as depicted on the left of .
	Without loss of generality we can assume that $x \in A$, because $int_x$ is an ancestor of $x$ before the $\nni$ move.
	If otherwise $x\in B$, we just need to exchange $A$ and $B$ in the following argumentation.

	If $A$ is a child of the lower of both new nodes after the $\nni$ move, as in the upper tree on the right of Figure~\ref{fig:nni_example_internal_label}, we can keep the rank order of $int_x$ and $int_y$.
	So we assign $int_x$ to the lower node that is parent of $A$ and $C$ and $int_y$ to the upper node that is parent of $B$.
	Because $x \in A$ and $y \in A \cup B \cup C$, $int_x$ and $int_y$ are ancestors of $x$ and $y$, respectively.

	Let us now consider the $\nni$ move that results in a tree where $B$ and $C$ are the children of the lower of the two new nodes and $A$ is children of the upper new node.
	Because $x \in A$, $int_x$ needs to be assigned to the upper new node.
	It remains to prove that $y \in B \cup C$ to make sure that $int_y$ is an ancestor of $y$ after the $\nni$ move.

	In the following we consider the tree after the $\nni$ move.
	We know that $y \in A \cup B \cup C$, because $int_y$ is an ancestor of $y$ before the $\nni$ move.
	Therefore, it is sufficient to prove $y \notin A$.
	Let us consider the subtree $A$ and the labels of its internal nodes.
	Let $|A| = k$ be the number of leaves in $A$.
	There are $k-1$ internal nodes in $A$ that need to be labelled by these leaves.
	The parent of the root of $A$ is $int_x$, as it is labelled by $x$.
	Because $x \in A$, there are $k - 1$ leaves remaining in $A$ that must label some internal nodes.
	As the $k-1$ internal nodes of $A$ must be labelled by leafs that are descendants of these nodes, these labels need to be the $k-1$ leafs of $A$.
	Therefore, it holds:
	$int_z \in A \iff z \in A$.
	And as we know that $int_y \notin A$, $y$ can not be a leaf of $A$.
	Therefore, it is $y \in B \cup C$ and thus, $int_y$ is an ancestor of $y$ after the $\nni$ move.

	Concluding we see that our labelling is well defined.
	Another thing we should point out is that this labelling method leads to the labelling of $T_2$ as it is defined for caterpillar trees.
	This follows directly from the fact that $int_z$ is an ancestor of the leaf labelled by $z$.
	If one considers the internal node in $T_2$ with lowest rank, it can only have the label of the leaf in the cherry that is not labelled by $a$, because $int_z$ is an ancestor of $z$ for all leaf labels $z$.
	Now we can stepwise consider all internal nodes with rank one higher than the previously considered node and receive the same result.
	Hence, our way to change the labels of internal nodes on $p$ is consistent and ends in the correct labelling of $T_2$.

	Now we want to use this labelling to prove that there is always a caterpillar path that is at least as short as the given shortest path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both, rank swaps and $\nni$ moves.
	Within a rank swap two nodes simply exchange their ranks.
	And within an $\nni$ move two new nodes arise and are labelled by the former two nodes that bounded the edge on that the $\nni$ move is performed.
	But still, the ranks of these two nodes $int_x$ and $int_y$ both change by $1$.

	On a caterpillar path the only possible moves are $\nni$ moves.
	However, each of these $\nni$ move change the ranks of two internal nodes by at most one each.
	Therefore, we can construct a caterpillar path $r$ out of $p$:
	every time two internla nodes $int_x$ and $int_y$ exchange ranks, they do so on the caterpillar path $r$ as well.
	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
\end{proof}

>>>>

%leaf or taxon?

@@

\begin{proof}
	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
	Let $p$ be a shortest path from $T_1$ to $T_2$.
	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.

@@

	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
	This makes sure that $int_x$ is an ancestor of $x$.
	Thus, it remains to prove that $int_y$ is ancestor of $y$.
	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
	Let us consider the tree before the $\nni$ move.
	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.

	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both rank swaps and $\nni$ moves, according to our definition of the labelling:
	Within a rank swap two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.

	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
	Therefore, we can construct a caterpillar path $r$ out of $p$:
	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
\end{proof}
